μGUIDE: a framework for quantitative imaging via generalized uncertainty-driven inference using deep learning

Version 1

Abstract

Abstract

This work proposes μGUIDE: a general Bayesian framework to estimate posterior distributions of tissue microstructure parameters from any given biophysical model or signal representation, with exemplar demonstration in diffusion-weighted MRI. Harnessing a new deep learning architecture for automatic signal feature selection combined with simulationbased inference and efficient sampling of the posterior distributions, μGUIDE bypasses the high computational and time cost of conventional Bayesian approaches and does not rely on acquisition constraints to define model-specific summary statistics. The obtained posterior distributions allow to highlight degeneracies present in the model definition and quantify the uncertainty and ambiguity of the estimated parameters.

1 Introduction

Diffusion-weighted Magnetic Resonance Imaging (dMRI) is a promising technique for characterizing brain microstructure in-vivo using a paradigm called microstructure imaging [Novikov et al., 2018a, Alexander et al., 2019, Jelescu et al., 2020]. Traditionally, microstructure imaging quantifies histologically meaningful features of brain microstructure by fitting a forward (biophysical) model voxel-wise to the set of signals obtained from images acquired with different sensitivities, yielding model parameter maps [Alexander et al., 2019].

Most commonly used techniques rely on a non-linear curve fitting of the signal and return the optimal solution, i.e. the best parameters guess of the fitting procedure. However, this may hide model degeneracy, that is all the other possible estimates that could explain the observed signal equally well [Jelescu et al., 2016]. Another crucial consideration in model fitting is accounting for the uncertainty in parameter estimates. This uncertainty serves various purposes, including assessing result confidence [Jones, 2003], quantifying noise effects [Behrens et al., 2003] or assisting in experimental design [Alexander, 2008].

Instead of attempting to remove the degeneracies, which has been the focus of a large number of studies [Palombo et al., 2023, De Almeida Martins et al., 2021, Slator et al., 2021, Jelescu et al., 2022, Warner et al., 2023, Uhl et al., 2024, Mougel et al., 2024, Olesen et al., 2022, Palombo et al., 2020, Howard et al., 2022, Jones et al., 2018, Vincent et al., 2020, Henriques et al., 2021, Afzali et al., 2021, Lampinen et al., 2023, Zhang et al., 2012-07, Guerreri et al., 2023, Gyori et al., 2022, Novikov et al., 2018b], we propose to highlight them and present all the possible parameter values that could explain an observed signal, providing users with more information to make more confident and explainable use of the inference results.

Posterior distributions are powerful tools to characterize all the possible parameter estimations that could explain an observed measurement, their uncertainty, and existing model degeneracy [Box and Tiao, 2011]. Bayesian inference allows for the estimation of these posterior distributions, traditionally approximating them using numerical methods, such as Markov-Chain-Monte-Carlo (MCMC) [Metropolis et al., 1953]. In quantitative MRI, these methods have been used for example to estimate brain connectivity [Behrens et al., 2003], optimize imaging protocols [Alexander, 2008], or infer crossing fibers by combining multiple spatial resolutions [Sotiropoulos et al., 2013]. However, these classical Bayesian inference methods are computationally expensive and time consuming. They also often require adjustments and tuning specific to each biophysical model [Harms and Roebroeck, 2018].

Harnessing a new deep learning architecture for automatic signal feature selection and efficient sampling of the posterior distributions using simulation-based inference [Cranmer et al., 2020, Lueckmann et al., 2017, Papamakarios et al., 2019], here we propose μGUIDE: a general Bayesian framework to estimate posterior distributions of tissue microstructure parameters from any given biophysical model/signal representation. μGUIDE extends and generalises previous work [Jallais et al., 2022] to any forward model and without acquisition constraints, providing fast estimations of posterior distributions voxel-wise. We demonstrate μGUIDE using numerical simulations on three biophysical models of increasing complexity and degeneracy and compare the obtained estimates with existing methods, including the classical MCMC approach. We then apply the proposed framework to dMRI data acquired from healthy human volunteers and participants with epilepsy. μGUIDE framework is agnostic to the origin of the data and the details of the forward model, so we envision its usage and utility to perform Bayesian inference of model parameters also using data from other MRI modalities (e.g. relaxation MRI) and beyond.

2 Results

2.1 Framework overview

The full architecture of the proposed Bayesian framework, dubbed μGUIDE, is presented in Fig. 1. μGUIDE allows to efficiently estimate full posterior distributions of tissue parameters. It is comprised of two modules that are optimized together to minimize the Kullback-Leibler divergence between the true posterior distribution and the estimated one for every parameters of a given forward model. The ‘Neural Posterior Estimator’ (NPE) module [Papamakarios et al., 2017] uses normalizing flows [Papamakarios et al., 2021] to approximate the posterior distribution, while the ‘Multi-Layer Perceptron’ (MLP) module is used to reduce the data dimensionality and ensure fast and robust convergence of the NPE module.

μGUIDE framework.

μGUIDE takes as input an observed data vector and relies on the definition of a biophysical or computational model [Ascoli et al., 2007, Callaghan et al., 2020, Jelescu et al., 2020]. It outputs a posterior distribution of the model parameters. Based on a SBI framework, it combines a Multi-Layer Perceptron (MLP) with 3 layers and a Neural Posterior Estimator (NPE). The MLP learns a low-dimensional representation of x, based on a small number of features (Nf), that can be either defined a priori or determined empirically during training. The MLP is trained simultaneously with the NPE, leading to the extraction of the optimal features that minimize the bias and uncertainty of p(θ|x).

The full posterior distribution contains a lot of useful information. To summarize and easily visualize this information, we propose three measures that quantify the best estimates and the associated confidence levels, and a way to highlight degeneracy. The three measures are the Maximum A Posteriori (MAP), which corresponds to the most likely parameter estimate; an uncertainty measure, which quantifies the dispersion of the 50% most probable samples using the interquartile range, relative to the prior range; and an ambiguity measure, which measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range. Fig. 2 presents those measures on exemplar posterior distributions.

μGUIDE summarizes information contained in the estimated posterior distributions.

A) Examples of degenerate and non-degenerate posterior distributions. B) Presentation of the measures introduced to quantify a posterior distribution on exemplar non-degenerate posterior distributions. Maximum A Posteriori (MAP): is the most likely parameter estimate (dashed vertical lines). Uncertainty: measures the dispersion of the 50% most probable samples using the interquartile range, with respect to the prior range. Ambiguity: measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range.

We show exemplar applications of μGUIDE to three biophysical models of increasing complexity and degeneracy from the dMRI literature: Ball&Stick [Behrens et al., 2003] (Model 1); Standard Model (SM) [Novikov et al., 2018a] (Model 2); and extended-SANDI [Palombo et al., 2020] (Model 3).

2.2 Evaluation of μGUIDE on simulations

2.2.1 Comparison with MCMC

We performed a comparison between the posterior distributions obtained using μGUIDE and MCMC, a classical Bayesian method. Fig. 3A shows posterior distributions on three exemplar simulations with SNR = 50 using the model 2 (SM), obtained with 15000 samples. Sharper and less biased posterior estimations are obtained using μGUIDE. Fig. 3B presents histograms for each model parameter of the bias between the ground truth value used to simulate a signal, and the MAP of the posterior distributions obtained with either μGUIDE or MCMC, on 200 simulations. Results indicate that the bias is similar or smaller using μGUIDE. Overall, μGUIDE posterior distributions are more accurate than the ones obtained with MCMC.

Comparison between μGUIDE and MCMC.

A) Posterior distributions obtained using either μGUIDE or MCMC on three exemplar simulations with model 2 (SM - SNR = 50). B) Bias between the ground truth values used for simulating the diffusion signals, and the maximum-a-posteriori extracted from the posterior distributions using either μGUIDE or MCMC. Sharper and less biased posterior distributions are obtained using μGUIDE.

Moreover, it took on average 29.3 s to obtain the posterior distribution using MCMC on a GPU (NVIDIA GeForce GT 710) for one voxel, while it only took 0.02 s for μGUIDE. μGUIDE is about 1500 times faster than MCMC, which makes it more suitable for applying it on large datasets.

2.2.2 The importance of feature selection

Fig. 4 shows the MAP extracted from the posterior distributions versus the ground truth parameters used to generate the diffusion signal with μGUIDE and manually-defined summary statistics for the three models. Less biased MAPs with lower ambiguities and uncertainties are obtained with μGUIDE, indicating that the MLP allows for the extraction of additional information not contained in the summary statistics, helping to solve the inverse problem with higher accuracy and precision. μGUIDE generalizes the method developed in [Jallais et al., 2022] to make it applicable to any forward model and any acquisition protocol, while making the estimates more precise and accurate thanks to the automatic feature extraction.

Fitting accuracy comparison between μGUIDE’s MLP-extracted features and manually-defined summary statistics.

Maximum-A-Posterioris extracted from the posterior distributions vs ground truth parameters used for generating the signal for the three models. Orange points correspond to the MAPs obtained using MLP-extracted features (μGUIDE) and the blue ones to the MAPs with the manually-defined summary statistics. Only the non-degenerate posterior distributions were kept. The summary statistics used in those three models are the direction-averaged signal for the Ball&Stick model, the LEMONADE system of equations [Novikov et al., 2018b] for the SM, and the summary statistics defined in [Jallais et al., 2022] for the extended-SANDI model. Results are shown on 100 exemplar noise-free simulations with random parameter combinations. The optimal features extracted by the MLP allow to reduce the bias and variance of the obtained microstructure posterior distributions.

2.2.3 μGUIDE highlights degeneracies

Fig. 5 presents the posterior distributions of microstructure parameters for the three models obtained with μGUIDE on exemplar noise-free simulations. Blue curves correspond to non-degenerate posterior distributions, while the red ones present at least one degeneracy for one of the parameters. As the complexity of the model increases, degeneracy in the model definitions appear. This figure showcases μGUIDE ability to highlight degeneracy in the model parameter estimation.

Exemplar posterior distributions of the microstructure parameters for the Ball&Stick, SM and extended SANDI models, obtained using μGUIDE on exemplar noise-free simulations.

As the complexity of the model increases, degeneracies appear (red posterior distributions). μGUIDE allows to highlight those degeneracies present in the model definition.

Tables 1 and 2 present the number of degenerate cases for each parameter in the three models, on 10000 simulations. Table 1 considers noise-free simulations and the training and estimations were performed on CPU. Table 2 reports results on noisy simulations (Rician noise with SNR = 50), with training and testing performed on a GPU (NVIDIA GeForce RTX 4090). The time needed for the inference and to estimate the posterior distributions on 10000 simulations, define if they are degenerate or not, and extract the MAP, uncertainty, ambiguity, are also reported. The more complex the model, the more degeneracies.

2.3 Application of μGUIDE to real data

After demonstrating that the proposed framework provides good estimates in the controlled case of simulations, we applied μGUIDE to both a healthy volunteer and a participant with epilepsy. The estimation of the posterior distributions is done independently for each voxel. To easily assess the values and the quality of the fitting, we are plotting the maximum-a-posterior, ambiguity and uncertainty maps, but the full posterior distributions are stored and available for all the voxels. Voxels presenting a degeneracy are highlighted with a red dot.

2.3.1 Healthy volunteer

We applied μGUIDE to a healthy volunteer, using the Ball&Stick, SM and extended-SANDI models. Fig. 6 presents the parametric maps of an exemplar set of model parameters for each model, alongside their degeneracy, uncertainty and ambiguity. The Ball&Stick model presents no degeneracy, the SM presents some degeneracy, mostly in voxels with high likelihood of partial voluming with cerebrospinal fluid (CSF) and at the white matter - gray matter boundaries. The extended-SANDI model is the model showing the highest number of degenerate cases, mostly localized within the white matter areas characterized by complex microstructure, e.g. crossing fibers. This result is expected, as the complexity of the models increases, leading to more combinations of tissue parameters that can explain an observed signal. Measures of ambiguity and uncertainty allow to quantify the confidence in the estimates and help interpreting the results.

Parametric maps of the Ball&Stick (top), SM (middle) and extended-SANDI model (bottom), obtained using μGUIDE.

Maximum-a-posteriori estimation, uncertainty and ambiguity measure maps are reported, overlayed with voxels considered degenerate (red dots).

2.3.2 Participant with epilepsy

Fig. 7 demonstrates μGUIDE application to a participant with epilepsy, using the SM. Noteworthy, the axonal signal fraction estimates within the epileptic lesion show low uncertainty and ambiguity measures hence high confidence, while orientation dispersion index estimates show high uncertainty and ambiguity suggesting low confidence, cautioning the interpretation.

Parametric maps of a participant with epilepsy obtained using μGUIDE with the SM, superimposed with the grey matter (black) and white matter (white) lesions segmentation.

Mean values of the maximum-a-posterior, uncertainty and ambiguity measures are reported in the two regions of interest. Lower MAP values are obtained in the lesions for the axonal signal fraction and the orientation dispersion index compared to healthy tissue. Higher uncertainty and ambiguity ODI values are reported, suggesting less stable estimations.

3 Discussion

3.1 Applicability of μGUIDE to multiple models

The μGUIDE framework offers the advantage of being easily applicable to various biophysical models and representations, thanks to its data-driven approach for data reduction. The need to manually define specific summary statistics that capture the relevant information for microstructure estimation from the multi-shell diffusion signal is removed. This also eliminates the acquisition constraints that were previously imposed by the summary statistics definition [Jallais et al., 2022]. The extracted features contain additional information compared to the summary statistics (see Appendix A), resulting in a notable reduction in bias (on average 5.2 fold lower), uncertainty (on average 2.6 fold lower) and ambiguity (on average 2.7 fold lower) in the estimated posterior distributions. Consequently, μGUIDE improves parameters estimation over current state-of-the-art methods (e.g. Jallais et al. [2022]), showing for example reduced bias (on average 5.2 fold lower) and dispersion (on average 6.4 fold lower) on the MAP estimates for each of the three example models investigated (see Fig. 4).

In this study, we presented applications of μGUIDE to brain microstructure estimation using three well-established biophysical models, with increased complexity: the Ball&Stick model, the Standard Model, and an extended-SANDI model. However, our approach is not limited to brain tissue nor to diffusion-weighted MRI and can be extended to different organs by employing their respective acquisition encoding and forward models, such as NEXI for exchange estimates [Jallais et al., 2024]; mcDESPOT for myelin water fraction mapping using quantitative MRI relaxation [Deoni et al., 2008]; VERDICT in prostate imaging [Panagiotaki et al., 2014]; or even adapted to different imaging modalities (e.g., EEG and MEG), where there is a way to link (via modelling or simulation) the observed signal to a set of parameters of interest. This versatility underscores the broad applicability of our proposed approach across various biological systems and imaging techniques.

It is important to note that μGUIDE is still a model-dependent method, meaning that the training process is based on the specific model being used. Additionally, the number of features extracted by the MLP needs to be predetermined. One way to determine the number of features is by matching it with the number of parameters being estimated. Alternatively, a dimensionality-reduction study using techniques like t-distributed stochastic neighbour embedding (tSNE) [Van der Maaten and Hinton, 2008] can be conducted to determine the optimal number of features.

3.2 μGUIDE: an efficient framework for Bayesian inference

One notable advantage of μGUIDE is its amortized nature. With this approach, the training process is performed only once, and thereafter, the posterior estimations can be independently obtained for all voxels. This amortization enables efficient estimations of the posterior distributions. μGUIDE outperforms in terms of speed conventional Bayesian inference methods such as MCMC, showing a “--1500 fold acceleration. The time savings achieved with μGUIDE make it a highly efficient and practical tool for estimating posterior distributions in a timely manner.

This unlocks the possibility to process with Bayesian inference very large datasets in manageable time (e.g. approximately 6 months to process 10k dMRI datasets) and to include Bayesian inference in iterative processes that require the repeated computation of the posterior distributions (e.g., dMRI acquisition optimization Alexander [2008]).

In the dMRI community, the use of SBI methods to characterize full posterior distributions as well as quantify the uncertainty in parameter estimations was first introduced in Jallais et al. [2022] for a grey matter model. An application to crossing fibers has recently been proposed by Karimi et al. [2024]. Those approaches use different density estimators. This work and Jallais et al. [2022] rely on Masked Autoregressive FLows (MAF) Papamakarios et al. [2017], while the work by Karimi et al. [2024] is based on Mixture Density Networks (MDN) [Bishop, 1994]. MAFs have been found to show superior performance compared to MDNs [Goncalves et al., 2020, Patron et al., 2022].

3.3 μGUIDE quantifies confidence to guide interpretation

Quantifying confidence in an estimate is of crucial importance. As demonstrated by our pathological example, changes in the tissue microstructure parameters can help clinicians decide which parameters are the most reliable and better interpret microstructure changes within diseased tissue. On large population studies, the quantified uncertainty can be taken into account when performing group statistics and to detect outliers.

Multiple approaches have been used to try and quantify this uncertainty. Gradient descent often provides a measure of confidence for each parameter estimate. Alternative approaches use the shape of the fitted tensor itself as a measure of uncertainty for the fiber direction [Koch et al., 2002, Parker and Alexander, 2003]. Other methods also rely on bootstrapping techniques to estimate uncertainty. Repetition bootstrapping for example depends on repeated measurements of signal for each gradient direction, but imply a long acquisition time and cost, and are prone to motion artifacts [Lazar and Alexander, 2005, Jones, 2003]. In contrast, residual bootstrapping methods resample the residuals of a regression model. Yet, this approach is heavily dependent on the model and can lead to overfitting [Whitcher et al., 2008, Chung et al., 2006]. In general, resampling methods can be problematic for sparse samples, as the bootstrapped samples tend to underestimate the true randomness of the distribution [Kauermann et al., 2009]. We propose to quantify the confidence by estimating full posterior distributions, which also has the benefit of highlighting degeneracy.

Variance observed in the posterior distributions can be attributed to several factors. The presence of noise in the signal contributes to irreducible variance, decreasing the confidence in the estimates as the noise level increases. Another source of variance can arise from the choice of acquisition parameters. Different acquisitions may provide varying levels of confidence in the parameter estimates. Under-sampled acquisitions or inadequate b-shells may fail to capture essential information about a tissue microstructure, such as soma or neurite radii, resulting in inaccurate estimates.

μGUIDE can guide users in determining whether an acquisition is suitable for estimating parameters of a given model and viceversa, the variance and bias of the posterior distributions estimated with μGUIDE can be used to guide the optimization of the data acquisition to maximize accuracy and precision of the model parameters estimates.

The presence of degeneracy in the solution of the inverse problem is influenced by the complexity of the model being used and the lack of sufficient information in the data. In recent years, researchers have introduced increasingly sophisticated models to better represent the brain tissue, such as SANDI [Palombo et al., 2020], NEXI [Jelescu et al., 2022] and eSANDIX [Olesen et al., 2022], that take into account an increasing number of tissue features. By applying μGUIDE, it becomes possible to gain insights into the degree of degeneracy within a model and to assess the balance between model realism and the ability to accurately invert the problem. We have recently provided an example of such application for NEXI and SANDIX [Jallais et al., 2024].

3.4 Summary

We propose a general Bayesian framework, dubbed μGUIDE, to efficiently estimate posterior distributions of tissue microstructure parameters. For any given acquisition and signal model/representation, μGUIDE improves parameters estimation and computational time over existing state-of-the-art methods. It allows to highlight degeneracy, and quantify confidence in the estimates, guiding results interpretation towards more confident and explainable diagnosis using modern deep learning. μGUIDE is not inherently limited to dMRI and microstructure imaging. We envision its usage and utility to perform efficient Bayesian inference also using data from any modality where there is a way to link (via modelling or simulation) the observed measurements to a set of parameters of interest.

4 Methods

4.1 Solving the inverse problem using Bayesian inference

4.1.1 The inference problem

We make the hypothesis that an observed dMRI signal x0 can be explained (and generated) using a handful of relevant tissue microstructure parameters θ0, following the definition of a forward model: x0=M(θ0)

The objective is, given this observation x0, to estimate the parameters ∞0 that generated it.

Forward models are designed to mimic at best a given biophysical phenomenon, for some given time and scale [Alexander, 2009, Yablonskiy and Sukstanskii, 2010, Jelescu and Budde, 2017, Novikov et al., 2018a, Alexander et al., 2019, Jelescu et al., 2020]. As a consequence, forward models are injection functions (every biologically plausible ∞i generates exactly one signal xi), but do not always happen to be bijections, meaning that multiple ∞i can generate the same signal xi. It can be impossible, based on biological considerations, to infer which solution ∞i best reflects the probed structure. We refer to these models as ‘degenerate models’.

Point estimates algorithms, such as minimum least square or maximum likelihood estimation algorithms, allow to estimate one set of microstructure parameters that could explain an observed signal. In the case of degenerate models, the solution space can be multi-modal and those algorithms will hide possible solutions. When considering real-life acquisitions, i.e. noisy and/or under-sampled acquisitions, one also needs to consider the bias introduced with respect to the forward model, and the resulting variance in the estimates [Jones, 2003, Behrens et al., 2003].

We propose a new framework that allows for the estimation of full posterior distributions p(∞|x0), that is all the probable parameters that could represent the underlying tissue, along with an uncertainty measure and the interdependency of parameters. These posteriors can help interpreting the obtained results and make more informed decisions.

4.1.2 The Bayesian formalism

The posterior distribution can be defined using Bayes’ theorem as follows: p(θ|x0)=p(x0|θ)p(θ)p(x0) (1)

where p(x0|θ) is the likelihood of the observed data point, p(θ) is the prior distribution defining our initial knowledge of the parameter values, and p(x0) is a normalizing constant, commonly referred to as the evidence of the data.

The evidence term is usually very hard to estimate, as it corresponds to all the possible realisations of x0, i.e. p(x0)=∫all x0p(x0|θ)p(θ) dx0 . For simplification, methods usually estimate an unnormalized probability density function, i.e. p(θ|x0) ∞ p(x0|θ)p(θ). (2)

To approximate these posterior distributions, traditional methods rely on the estimation of the likelihood p(x0|θ) of the observed data point x0 via an analytic expression. This likelihood function corresponds to an integral over all possible trajectories through the latent space, that is p(x0|θ) = ∫ p(x0, z|∞)dz, where p(x0, z|∞) is the joint probability density of observed data x0 and latent variables z. For forward models with large latent spaces, computing this integral explicitly becomes impractical. The likelihood function is then intractable, rendering these methods unusable [Cranmer et al., 2020]. Models that do not admit a tractable likelihood are called implicit models [Diggle and Gratton, 1984].

To circumvent this issue, some techniques have been proposed to sample numerically from the likelihood function, such as Markov-Chain-Monte-Carlo (MCMC) [Metropolis et al., 1953]. Another set of approaches proposes to train a conditional density estimator to learn a surrogate of the likelihood distribution [Papamakarios et al., 2019, Lueckmann et al., 2019], the likelihoodratio [Cranmer et al., 2016, Gutmann et al., 2018] or the posterior distribution [Papamakarios and Murray, 2016, Lueckmann et al., 2017, Papamakarios et al., 2019], allowing to greatly reduce computation times. These methods are dubbed Likelihood-Free Inference (LFI), or SimulationBased Inference (SBI) methods [Cranmer et al., 2020, Tejero-Cantero et al., 2020]. In particular, there has been a growing interest towards deep generative modeling approaches in the machine learning community [Lueckmann et al., 2021]. They rely on specially tailored neural network architectures to approximate probability density functions from a set of examples. Normalizing flows [Papamakarios et al., 2021] are a particular class of such neural networks that have demonstrated promising results for SBI in different research fields [Goncalves et al., 2020, Greenberg et al., 2019].

While this work focuses on the estimate of the posterior distribution using a conditional density estimator, we show a comparison with MCMC, which are commonly used methods in the community. We will therefore introduce this method in the following paragraph.

4.1.3 Estimating the likelihood function

Well-established approaches for estimating the likelihood function are MCMC methods. These methods rely on a noise model to define the likelihood distribution, such as the Rician [Panagiotaki et al., 2012] or Offset Gaussian models [Alexander, 2009]. In this work, we will be using the Microstructure Diffusion Toolbox to perform the MCMC computations [Harms and Roebroeck, 2018], which relies on the Offset Gaussian model. The log-likelihood function is then the following: log(p(x0|θ))=−∑(x0−M(θ)2+σ2)2σ2−m.log(σ2π) (3)

where M(∞) is the signal obtained using the biophysical model, σ is the standard deviation of the Gaussian distributed noise, estimated from the reconstructed magnitude images [Dietrich et al., 2007], and m is the number of observations in the dataset.

MCMC methods allow to obtain posterior distributions using Bayes’ formula (Eq. (2)) with the previously defined likelihood function (Eq. (3)) and some prior distributions, which are usually uniform distributions defined on biologically plausible ranges. They generate a multi-dimensional chain of samples which is guaranteed to converge towards a stationary distribution, which approximates the posterior distribution [Metropolis et al., 1953].

The need to compute the signal following the forward model at each iteration makes these sampling methods computationally expensive and time consuming. In addition, they require some adjustments specific to each model, such as the choice of burn-in length, thinning and the number of samples to store. Harms and Roebroeck [2018] recommend to use the Adaptive Metropolis-Within-Gibbs (AMWG) algorithm for sampling dMRI models, initialized with a maximum likelihood estimator obtained from non-linear optimization, with 100 to 200 samples for burn-in and no thinning. The recommended number of samples is model-dependent. Authors recommendations can be found in their paper.

4.1.4 Bypassing the likelihood function

An alternative method was proposed to overcome the challenges associated with approximating the likelihood function and the limitations of MCMC sampling algorithms. This approach involves directly approximating the posterior distribution by using a conditional density estimator, i.e. a family of conditional probability density function approximators denoted as qϕ(∞|x). These approximators are parameterized by ϕ and accept both the parameters θ and the observation x as input arguments. Our posterior approximation is then obtained by minimizing its average Kullback-Leibler divergence with respect to the conditional density estimator for different choices of x, as per [Papamakarios and Murray, 2016]: minϕ. L(ϕ) with L(ϕ)=Ex~p(x)[DKL(p(θ|x)||qϕ(θ|x))], (4)

which can be rewritten as L(ϕ)=∫DKL(p(θ|x)||qϕ(θ|x))p(x)dx,=−∫∫log(qϕ(θ|x))p(θ|x)p(x)dθdx+C,=−∫∫log(qϕ(θ|x))p(x,θ)dθdx+C,=−E(x,θ)~p(x,θ)[ log(qϕ(θ|x)) ]+C, (5)

where C is a constant that does not depend on ∞. Note that in practice we consider a N-sample Monte-Carlo approximation of the loss function: L(ϕ)≈LN(ϕ)=−1N∑i=1Nlog(qϕ(θi+xi)), (6)

where the N data points (θi, xi) are sampled from the joint distribution with θi ~ p(θ) and xi ~ p(x|θi). We can then use stochastic gradient descent to obtain a set of parameters ϕ which minimizes LN.

If the class of conditional density estimators is sufficiently expressive, it can be demonstrated that the minimizer of Eq. (6) converges to p(θ|y) when N → ∞ [Greenberg et al., 2019]. It is worth noting that the parametrization ϕ, obtained at the end of the optimization procedure, serves as an amortized posterior for various choices of x. Hence, for a particular observation x0, we can simply use qϕ(θ|x0) as an approximation of p(θ|x0).

4.2 μGUIDE framework

The full architecture of the proposed Bayesian framework, dubbed μGUIDE, is presented in Fig. 1. The analysis codes underpinning the results presented here can be found upon publication in the Cardiff University data catalogue and on Github: https://github.com/mjallais/uGUIDE (both CPU and GPU are supported).

μGUIDE is comprised of two modules that are optimized together to minimize the Kull-back-Leibler divergence between the true posterior distribution and the estimated one for every parameters of a given forward model. The ‘Neural Posterior Estimator’ (NPE) module uses normalizing flows to approximate the posterior distribution, while the ‘Multi-Layer Perceptron’ (MLP) module is used to reduce the data dimensionality and ensure fast and robust convergence of the NPE module. The following Sections provide more details about our implementation of each module.

4.2.1 Neural Posterior Estimator

In this study, the Sequential Neural Posterior Estimation (SNPE-C) algorithm [Papamakarios and Murray, 2016, Greenberg et al., 2019] with a single round is employed to train a neural network that directly approximates the posterior distribution. Thus, sampling from the posterior can be done by sampling from the trained neural network. Neural density estimators have the advantage of providing exact density evaluations, in contrast to Variational Autoencoders (VAE) Kingma and Welling [2019] or generative adversarial networks (GAN) Goodfellow et al. [2014], which are better suited for generating synthetic data.

The conditional probability density function approximators used in this pro ject belong to a class of neural networks called normalizing flows [Papamakarios et al., 2021]. These flows are invertible functions capable of transforming vectors generated from a simple base distribution (e.g. the standard multivariate Gaussian distribution) into an approximation of the true posterior distribution. An autoregressive architecture for normalizing flows is employed, implemented via the Masked Autoregressive Flow (MAF) Papamakarios et al. [2017], which is constructed by stacking five Masked Autoencoder for Distribution Estimation models (MADE) Germain et al. [2015]. An explanation of how MAF and MADE work is provided in Appendix B.

To test that the predicted posteriors for a given model are not incorrect we use posterior predictive checks, which is described in more details in Appendix C.

4.2.2 Handling the large dimensionality of the data with Multi-Layer Perceptron

As the dimensionality of the input data x grows, the complexity of the corresponding inverse problem also increases. Accurately characterizing the posterior distributions or estimating the tissue microstructure parameters becomes more challenging. As a consequence, it is often necessary to rely on a set of low-dimensional features (or summary statistics) instead of the raw data for the inference task process [Blum et al., 2013, Fearnhead and Prangle, 2012, Papamakarios et al., 2019]. These summary statistics are features that capture the essential information within the raw data, allowing to reduce the size of the input vector. Learning a set of sufficient statistics before estimating the posterior distribution makes the inference easier and offers many benefits (see e.g. the Rao-Blackwell theorem).

A follow-up challenge lies in the choice of suitable summary statistics. For well-understood problems and data, it is possible to manually design these features using deterministic functions that condense the information contained in the raw signal into a set of handful summary statistics. Previous works, such as Novikov et al. [2018b] and Jallais et al. [2022], have proposed specific summary statistics for two different biophysical models. However, defining these summary statistics is difficult and often requires prior knowledge of the problem at hand. In the context of dMRI, they also rely on acquisition constraints and are model-specific.

In this work, the proposed framework aims to be applicable to any forward model and be as general as possible. We therefore propose to learn the summary statistics from the high-dimensional input signals x using a neural network. This neural network is referred to as an embedding neural network. The observed signals are fed into the embedding neural network, whose outputs are then passed to the neural density estimator. The parameters of the embedding network are learned together with the parameters of the neural density estimator, leading to the extraction of optimal features that minimize the uncertainty of p(ϕ|x). Here, we propose to use a MLP with three layers as a summary statistics extractor. The number of features Nf extracted by the MLP can be either defined a priori or determined empirically during training.

4.2.3 Training μGUIDE

To train μGUIDE we need couples of input vectors x and corresponding ground-truth values for the model parameters that we want to estimate, θ. The input x can be real or simulated data (e.g. dMRI signals); or a mixture of these two. We train μGUIDE by stochastically minimizing the loss function defined in Eq. (6) using the Adam optimizer [Kingma and Ba, 2015] with a learning rate of 10−3 and a minibatch size of 128. We use 1 million simulations for each model, 5% of which are randomly selected to be used as a validation set. Training is stopped when the validation loss does not decrease for 30 consecutive epochs.

4.2.4 Quantifying the confidence in the estimates

The full posterior distribution contains a lot of useful information about a given model parameter best estimates; uncertainty; ambiguity and degeneracy. To summarize and easily visualize this information, we propose three measures that quantify the best estimates and the associated confidence levels, and a way to highlight degeneracy.

We start by checking whether a posterior distribution is degenerate, that is if the distribution presents multiple distinct parameter solutions, appearing as multiple local maxima (Fig. 2). To that aim, we fit two Gaussian distributions to the obtained posterior distributions. A voxel is considered as degenerate if the derivative of the fitted Gaussian distributions changes signs more than once (i.e. multiple local maxima), and if the two Gaussian distributions are not overlapping (the distance between the two Gaussian means is inferior to the sum of their standard deviations).

For non-degenerate posterior distributions, we extract three quantities:

The Maximum A Posteriori (MAP), which corresponds to the most likely parameter estimate.

An uncertainty measure, which quantifies the dispersion of the 50% most probable samples using the interquartile range, relative to the prior range.

An ambiguity measure, which measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range.

Fig. 2 presents those measures on exemplar posterior distributions.

4.3 Application of μGUIDE to biophysical modelling of dMRI data

We show exemplar applications of μGUIDE to three biophysical models of increasing complexity and degeneracy from the dMRI literature. For each model, we compare the fitting quality of the posterior distributions obtained using the MLP and manually defined summary statistics.

4.3.1 Biophysical models of dMRI signal.

Model 1: Ball&Stick [Behrens et al., 2003]. This is a two-compartment model (intra-neurite and extra-neurite space) where the dMRI signal from the brain tissue is modeled as a weighted sum, with weight fin, of signals from water diffusing inside the neurites, approximated as sticks (i.e. cylinders of zero radius) with diffusivity Din, and water diffusing within the extra-neurite space, approximated as Gaussian diffusion in an isotropic medium with diffusivity De. The direction of the stick is randomly sampled on a sphere. This model has the main advantage of being nondegenerate. We define the summary statistics as the direction-averaged signal (six b-shells, see Section 4.3.3).

Model 2: Standard Model (SM) [Novikov et al., 2018a]. Expanding on model 1, this model represents the dMRI signal from the brain tissue as a weighted sum of the signal from water diffusing within the neurite space, approximated as sticks with symmetric orientation dispersion following a Watson distribution and water diffusing within the extra-neurite space, modelled as anisotropic Gaussian diffusion. The microstructure parameters of this two-compartment model are the neurite signal fraction f , the intra-neurite diffusivity Da, the orientation dispersion index ODI, and the parallel and perpendicular diffusivities within the extra-neurite space De|| and De⊥ We use the LEMONADE [Novikov et al., 2018b] system of equations, which is based on a cumulant decomposition of the signal, to define six summary statistics.

Model 3: extended-SANDI [Palombo et al., 2020]. This is a three-compartment model (intra-neurite, intra-soma and extra-cellular space) where the dMRI signal from the brain tissue is modelled as a weighted sum of the signal from water diffusing within the neurite space, approximated as sticks with symmetric orientation dispersion following a Watson distribution; water diffusing within cell bodies (namely soma), modelled as restricted diffusion in spheres; and water diffusing within the extra-cellular space, modelled as isotropic Gaussian diffusion. The parameters of interest are the neurite signal fraction fn, the intra-neurite diffusivity Dn, the orientation dispersion index ODI, the extra-cellular signal fraction fe and isotropic diffusivity De, the soma signal fraction fs, and a proxy of soma radius and diffusivity Cs, defined as [Jallais et al., 2022]: Cs2Dsδ2∑αm2rs2−2∞αm−4αm2rs2−2⋅(2δ−2+e−αm2Ds(Δ−δ)−e−αm2Dsδ−e−αm2DsΔ+e−αm2Ds(Δ+δ)αm2Ds),

with rs and Ds the soma radius and diffusivity respectively, and αm the mth root of (αrs)−1J32(αrs)=J52(αrs) , with Jn(x) the Bessel functions of the first kind. We use the six summary statistics defined in [Jallais et al., 2022], which are based on a high and low b-value signal expansion.

Prior distributions p(θ) are defined as uniform distributions over biophysically plausible ranges. Signal fractions are defined within the interval [0,1], diffusivities between 0.1 and 3 μm2 ms-1, ODI between 0.03 and 0.95, and Cs between 0.15 and 1105 μm2 (which correspond to rs ∈ [1; 15] μm and fixed Ds = 3 μm2 ms-1).

The SM imposes the constraint De⊥ < De|| To generate samples uniformly distributed on the space defined by this condition, we are using two random variables u0 and u1, both sampled uniformly between 0 and 1, and then relate them to De|| and De⊥ using the following equations: { De||=(3.0−0.1)2⋅u0+0.1De⊥=(De||−0.1)⋅u10.1 (7)

The extended-SANDI model requires for the signal fractions to sum to 1, that is fn+fs+fk = 1. To uniformly cover the simplex fn + fs + fe = 1, we define two new parameters k1 and k2 , uniformly sampled between 0 and 1, and use the following equations to get the corresponding signal fractions: { fn=k2k1fs=(1−k2)k1fe=1−k1 (8)

To ensure comparability of results, we extract the same number of features Nf using the MLP as the number of summary statistics for each model. We therefore use Nf = 6 for the Ball&Stick, the SM and the extended-SANDI models.

4.3.2 Validation in Numerical Simulations

We start by validating the proposed method using posterior predictive checks (PPC) and simulated signals from model 2 (see more details in Appendix C). Since PPC alone does not guarantee the correctness of the estimated posteriors, we further validated the obtained posterior distributions comparing them with the Adaptive Metropolis-Within-Gibbs MCMC [Roberts and Rosenthal, 2009]. We generated simulations following the same acquisition protocol as the real data (see Section 4.3.3), and added Rician distributed noise with a Signal-to-Noise-Ratio (SNR) of 50. We then estimated the posterior distributions using both μGUIDE and the MCMC method implemented in the MDT toolbox [Harms and Roebroeck, 2018]. We initialized the sampling using a maximum likelihood estimator. We sampled 15200 samples from the distribution, the first 200 ones being used as burn-in, and no thinning. Similarly, we sampled 15000 samples from the estimated posterior distributions using μGUIDE.

Then, we show that μGUIDE can be applicable to any model. We use model 1 and 3 as examples of simpler (and non-degenerate) and more complex (and degenerate) models than model 2, respectively.

We compared the proposed framework to a state-of-the-art method for posterior estimation [Jallais et al., 2022]. This method relies on manually-defined summary statistics, while μGUIDE automatically extracts some features using an embedded neural network. μGUIDE was trained directly on noisy simulations. The manually-defined summary statistics were extracted from these simulated noisy signals and then used as training dataset for a MAF, similarly to [Jallais et al., 2022].

Finally, we used μGUIDE to highlight degeneracy in all the models. While the complexity of the models increases, more degeneracy can be found. The degeneracy is inherent to the model definition, and is not induced by the noise. μGUIDE allows to highlight those degeneracies and quantify the confidence in the obtained estimates.

The training was performed on N = 106 numerical simulations for each model, computed using the MISST package [lanus et al., 2017] and random combinations of the model parameters, each uniformly sampled from the previously defined ranges, with the addition of Rician distributed noise with SNR equivalent to the experimental data, i.e. 50.

4.3.3 dMRI Data Acquisition and Processing

We applied μGUIDE to dMRI data collected from two participants: a healthy volunteer from the WAND dataset [McNabb et al., 2024] and an age-matched participant with epilepsy, acquired with the same protocol used for the MICRA dataset [Koller et al., 2021]. Data were acquired on a Connectome 3T scanner using a single-shot spin-echo, echo-planar imaging sequence with b-values = [200, 500, 1200, 2400, 4000, 6000] s mm-2, [20, 20, 30, 61, 61, 61] uniformly distributed directions respectively and 13 non-diffusion-weighted images at 2 mm isotropic resolution. TR was set to 3000 ms, TE to 59 ms, and the diffusion gradient duration and separation to 7 ms and 24 ms respectively. Short diffusion times and TE were achieved thanks to the Connectom gradients, allowing to enhance the signal-to-noise ratio and sensitivity to small water displacements [Jones et al., 2018, Setsompop et al., 2013]. We considered the noise as Rician with an SNR of 50 for both subjects.

Data were preprocessed using a combination of in-house pipelines and tools from the FSL [Andersson et al., 2003, Andersson and Sotiropoulos, 2016, Smith, 2002, Smith et al., 2004] and MRTrix3 [Tournier et al., 2019] software packages. The preprocessing steps included brain extraction [Smith, 2002], denoising [Cordero-Grande et al., 2019, Veraart et al., 2016], drift correction [Vos et al., 2017, Sairanen et al., 2018], susceptibility-induced distortions [Andersson et al., 2003, Smith et al., 2004], motion and eddy current correction [Andersson and Sotiropoulos, 2016], correction for gradient non-linearity distortions [Glasser et al., 2013], and Gibbs ringing artefacts correction [Kellner et al., 2016].

4.3.4. dMRI Data Analysis

Diffusion signals were first normalized by the mean non-diffusion-weighted signals acquired for each voxel. Each voxel was then estimated in parallel using the μGUIDE framework. For each observed signal xv (i.e. for each voxel), we drew 50000 samples via rejection-sampling from qϕ(θi|xv) for each model parameter θi, allowing to retrieve the full posterior distributions. If a posterior distribution was not deemed degenerate, the maximum-a-posteriori, uncertainty and ambiguity measures were extracted from the posterior distributions.

The manually-defined summary statistics of the SM are defined using a cumulant expansion, which is only valid for small b-values. We therefore only used the b ≤ 2500 smm-2 data for this model. In order to obtain comparable results, we restricted the application of μGUIDE to this range of b-values as well. An extra b-shell (b-value = 5000 s mm-2; 61 directions) was interpolated using mapl [Fick et al., 2016] for the extended-SANDI model when using the method developed by Jallais et al. [2022] based on summary statistics.

The training of μGUIDE was performed as described in 4.3.2 and an example of training dataset and input signal vector is provided in Fig. 8.

Example training set and input signals for μGUIDE.

A) Examples of input synthetic data vectors and corresponding ground truth model parameters used in the training set of Model 1 (Ball&Stick). B) Example of input measured signals from a voxel in a healthy participant, used for inference.

All the computations were performed both on CPU and GPU (NVIDIA GeForce RTX 4090).

eLife Assessment

The authors proposed a novel deep learning framework to estimate posterior distributions of tissue microstructure parameters. This provides a valuable methodology with practical implications for automatically estimating parameter distributions from different biophysical models. The experiments show solid evidence for generalizing the method to use data from different protocol acquisitions and work with models of varying complexity.

Reviewer #1 (Public Review):

The authors proposed a framework to estimate the posterior distribution of parameters in biophysical models. The framework has two modules: the first MLP module is used to reduce data dimensionality and the second NPE module is used to approximate the desired posterior distribution. The results show that the MLP module can capture additional information compared to manually defined summary statistics. By using the NPE module, the repetitive evaluation of the forward model is avoided, thus making the framework computationally efficient. The results show the framework has promise in identifying degeneracy. This is an interesting work.

Reviewer #2 (Public Review):

Summary:

The authors improve the work of Jallais et al. (2022) by including a novel module capable of automatically learning feature selection from different acquisition protocols inside a supervised learning framework. Combining the module above with an estimation framework for estimating the posterior distribution of model parameters, they obtain rich probabilistic information (uncertainty and degeneracy) on the parameters in a reasonable computation time.

The main contributions of the work are:

(1) The whole framework allows the user to avoid manually defining summary statistics, which may be slow and tedious and affect the quality of the results.

(2) The authors tested the proposal by tackling three different biophysical models for brain tissue and using data with characteristics commonly used by the diffusion-MR-microstructure research community.

(3) The authors validated their method well with the state-of-the-art.

The main weakness is:

(1) The methodology was tested only on scenarios with a signal-to-noise ratio (SNR) equal to 50. It is interesting to show results with lower SNR and without noise that the method can detect the model's inherent degenerations and how the degeneration increases when strong noise is present. I suggest expanding the Figure in Appendix 1 to include this information.

The authors showed the utility of their proposal by computing complex parameter descriptors automatically in an achievable time for three different and relevant biophysical models.

Importantly, this proposal promotes tackling, analyzing, and considering the degenerated nature of the most used models in brain microstructure estimation.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

The authors proposed a framework to estimate the posterior distribution of parameters in biophysical models. The framework has two modules: the first MLP module is used to reduce data dimensionality and the second NPE module is used to approximate the desired posterior distribution. The results show that the MLP module can capture additional information compared to manually defined summary statistics. By using the NPE module, the repetitive evaluation of the forward model is avoided, thus making the framework computationally efficient. The results show the framework has promise in identifying degeneracy. This is an interesting work.

We thank the reviewer for the positive comments made on our manuscript.

Reviewer #1 (Recommendations For The Authors):

I have some minor comments.

(1) The uGUIDE framework has two modules, MLP and NPE. Why are the two modules trained jointly? The MLP module is used to reduce data dimensionality. Given that the number of features for different models is all fixed to 6, why does one need different MLPs? This module should, in principle, be general-purpose and independent of the model used.

The MLP must be trained together with the NPE module to maximise inference performance in terms of accuracy and precision. Although the number of features predicted by the MLP was fixed to six, the characteristics of these six features can be very different, depending on the chosen forward model and the available data, as we showed in Appendix 1 Figure 1. Training the MLP independently of the NPE would result in suboptimal performance of µGUIDE, with potentially higher bias and variance of the predicted posterior distributions. We have now added these considerations in the Methods section.

(2) The authors mentioned at L463 that all the 3 models use 6 features. From L445 to L447, it seems model 3 has 7 unknown parameters. How can one use 6 features to estimate 7 unknowns?

Thank you for pointing out the lack of clarity regarding the parameters to estimate in this section. Model 3 is a three-compartment model, whose parameters of interest are the signal fraction and diffusivity from water diffusing in the neurite space (fn and Dn), the neurites orientation dispersion index (ODI), the signal fraction in cell bodies (fs), a proxy to soma radius and diffusivity (Cs), and the signal fraction and diffusivity in the extracellular space (fe and De). The signal fractions are constrained by the relationship fn + fs + fe = 1, hence _fe i_s calculated from the estimated fn and fs. This leaves us with 6 parameters to estimate: fn, Dn, ODI, fs, Cs, De. We clarified it in the revised version of the paper.

(3) L471, Rician noise is not a proper term. Rician distribution is the distribution of pixel intensities observed in the presence of noise. And Rician distribution is the result of magnitude reconstruction. See "Noise in magnitude magnetic resonance images" published in 2008. I assume that real-valued Gaussian noise is added to simulated data.

We apologize for the confusion. We added Gaussian noise to the real and imaginary parts of the simulated signals and then used the magnitude of this noisy complex signal for our experiments. We rephrased the sentence for more clarity.

(4) L475, why thinning is not used in MCMC? In figure 3, the MCMC results are more biased than uGUIDE, is it related to no thinning in MCMC?

We followed the recommendations by Harms et al. (2018) for the MCMC experiments. They analysed the impact of thinning (among other parameters) on the estimated posterior distributions. Their findings indicate that thinning is unnecessary and inefficient, and they recommend using more samples instead. For further details, we refer the reviewer to their publication, along with the theoretical works they cite. We have now added this note in the Methods section.

(5) Did the authors try model-fitting methods with different initializations to get a distribution of the parameters? Like the paper "Degeneracy in model parameter estimation for multi‐compartmental diffusion in neuronal tissue". For the in vivo data, it is informative to see the model-fitting results.

No, we did not try model-fitting methods with different initializations because such methods provide only a partial description of the solution landscape, which can be interpreted as a partial posterior distribution. Although this approach can help to highlight the problem of degeneracy, it does not provide a complete description of all potential solutions. In contrast, MCMC estimates the full posterior distribution, offering a more accurate and precise characterization of degeneracies and uncertainties compared to model-fitting methods with varying initializations. Hence, we decided to use MCMC as benchmark. We have now added these considerations to the Discussion section.

Reviewer #2 (Public Review):

Summary:

The authors improve the work of Jallais et al. (2022) by including a novel module capable of automatically learning feature selection from different acquisition protocols inside a supervised learning framework. Combining the module above with an estimation framework for estimating the posterior distribution of model parameters, they obtain rich probabilistic information (uncertainty and degeneracy) on the parameters in a reasonable computation time.

The main contributions of the work are:

(1) The whole framework allows the user to avoid manually defining summary statistics, which may be slow and tedious and affect the quality of the results.

(2) The authors tested the proposal by tackling three different biophysical models for brain tissue and using data with characteristics commonly used by the diffusion-MRmicrostructure research community.

(3) The authors validated their method well with the state-of-the-art.

The main weakness is:

(1) The methodology was tested only on scenarios with a signal-to-noise ratio (SNR) equal to 50. It is interesting to show results with lower SNR and without noise that the method can detect the model's inherent degenerations and how the degeneration increases when strong noise is present. I suggest expanding the Figure in Appendix 1 to include this information.

The authors showed the utility of their proposal by computing complex parameter descriptors automatically in an achievable time for three different and relevant biophysical models.

Importantly, this proposal promotes tackling, analysing, and considering the degenerated nature of the most used models in brain microstructure estimation.

We thank the reviewer for these positive remarks.

Concerning the main weakness highlighted by the reviewer: In our submitted work, we presented results both without noise and with a signal-to-noise ratio (SNR) equal to 50 (similar to the SNR in the experimental data analysed). Figure 5 shows exemplar posterior distributions obtained in a noise-free scenario, and Table 1 reports the number of degeneracies for each model on 10000 noise-free simulations. These results highlight that the presence of degeneracies is inherent to the model definition. Figures 3, 6 and 7 present results considering an SNR of 50. We acknowledge that results with lower SNR have not been included in the initial submission. To address this, we added a figure in the appendix illustrating the impact of noise on the posterior distributions. Specifically, Figure 1A of Appendix 2 shows posterior distributions estimated from signals generated using an exemplar set of model parameters with varying noise levels

(no noise, SNR=50 and SNR=25). Figure 1B presents uncertainties values obtained on 1000 simulations for each noise level. We observe that, as the SNR reduces, uncertainty increases. Noise in the signal contributes to irreducible variance. The confidence in the estimates therefore reduces as the noise level increases.

Reviewer #2 (Recommendations For The Authors):

Some suggestions:

Panel A of Figure 2 may deserve a better explanation in the Figure's caption.

We agree that the description of panel A of figure 2 was succinct and added more explanation in the figure’s caption.

The caption of Figure 3 should mention that the panel's titles are the parameters of the used biophysical models.

We added in the caption of figure 3 that the names of the model parameters are indicated in the titles of the panels. We apologise for the confusion it may have created.

In equation (3), the authors should indicate the summation index.

We apologise for not putting the summation index in equation 3. We added it in the revised version.

In line 474, the authors should discuss if the systematic use of the maximum likelihood estimator as an initializer for the sampling does not bias the computed results.

Concerning the MCMC estimations, we followed the recommendations from Harms et al. (2018). They investigated the use of starting from the maximum likelihood estimator (MLE). They concluded that starting from the MLE allows to start in the stationary distribution of the Markov chain, removing the need for some burn-in. Additionally, they showed that initializing the sampling from the MLE has the advantage of removing salt- and pepper-like noise from the resulting mean and standard deviation maps. We have now added this note in the Methods section.

Version 2

Abstract

Abstract

This work proposes μGUIDE: a general Bayesian framework to estimate posterior distributions of tissue microstructure parameters from any given biophysical model or signal representation, with exemplar demonstration in diffusion-weighted MRI. Harnessing a new deep learning architecture for automatic signal feature selection combined with simulationbased inference and efficient sampling of the posterior distributions, μGUIDE bypasses the high computational and time cost of conventional Bayesian approaches and does not rely on acquisition constraints to define model-specific summary statistics. The obtained posterior distributions allow to highlight degeneracies present in the model definition and quantify the uncertainty and ambiguity of the estimated parameters.

1 Introduction

Diffusion-weighted Magnetic Resonance Imaging (dMRI) is a promising technique for characterizing brain microstructure in-vivo using a paradigm called microstructure imaging [Novikov et al., 2018a, Alexander et al., 2019, Jelescu et al., 2020]. Traditionally, microstructure imaging quantifies histologically meaningful features of brain microstructure by fitting a forward (biophysical) model voxel-wise to the set of signals obtained from images acquired with different sensitivities, yielding model parameter maps [Alexander et al., 2019].

Most commonly used techniques rely on a non-linear curve fitting of the signal and return the optimal solution, i.e. the best parameters guess of the fitting procedure. However, this may hide model degeneracy, that is all the other possible estimates that could explain the observed signal equally well [Jelescu et al., 2016]. Another crucial consideration in model fitting is accounting for the uncertainty in parameter estimates. This uncertainty serves various purposes, including assessing result confidence [Jones, 2003], quantifying noise effects [Behrens et al., 2003] or assisting in experimental design [Alexander, 2008].

Instead of attempting to remove the degeneracies, which has been the focus of a large number of studies [Palombo et al., 2023, De Almeida Martins et al., 2021, Slator et al., 2021, Jelescu et al., 2022, Warner et al., 2023, Uhl et al., 2024, Mougel et al., 2024, Olesen et al., 2022, Palombo et al., 2020, Howard et al., 2022, Jones et al., 2018, Vincent et al., 2020, Henriques et al., 2021, Afzali et al., 2021, Lampinen et al., 2023, Zhang et al., 2012-07, Guerreri et al., 2023, Gyori et al., 2022, Novikov et al., 2018b], we propose to highlight them and present all the possible parameter values that could explain an observed signal, providing users with more information to make more confident and explainable use of the inference results.

Posterior distributions are powerful tools to characterize all the possible parameter estimations that could explain an observed measurement, their uncertainty, and existing model degeneracy [Box and Tiao, 2011]. Bayesian inference allows for the estimation of these posterior distributions, traditionally approximating them using numerical methods, such as Markov-Chain-Monte-Carlo (MCMC) [Metropolis et al., 1953]. In quantitative MRI, these methods have been used for example to estimate brain connectivity [Behrens et al., 2003], optimize imaging protocols [Alexander, 2008], or infer crossing fibers by combining multiple spatial resolutions [Sotiropoulos et al., 2013]. However, these classical Bayesian inference methods are computationally expensive and time consuming. They also often require adjustments and tuning specific to each biophysical model [Harms and Roebroeck, 2018].

Harnessing a new deep learning architecture for automatic signal feature selection and efficient sampling of the posterior distributions using simulation-based inference [Cranmer et al., 2020, Lueckmann et al., 2017, Papamakarios et al., 2019], here we propose μGUIDE: a general Bayesian framework to estimate posterior distributions of tissue microstructure parameters from any given biophysical model/signal representation. μGUIDE extends and generalises previous work [Jallais et al., 2022] to any forward model and without acquisition constraints, providing fast estimations of posterior distributions voxel-wise. We demonstrate μGUIDE using numerical simulations on three biophysical models of increasing complexity and degeneracy and compare the obtained estimates with existing methods, including the classical MCMC approach. We then apply the proposed framework to dMRI data acquired from healthy human volunteers and participants with epilepsy. μGUIDE framework is agnostic to the origin of the data and the details of the forward model, so we envision its usage and utility to perform Bayesian inference of model parameters also using data from other MRI modalities (e.g. relaxation MRI) and beyond.

2 Results

2.1 Framework overview

The full architecture of the proposed Bayesian framework, dubbed μGUIDE, is presented in Fig. 1. μGUIDE allows to efficiently estimate full posterior distributions of tissue parameters. It is comprised of two modules that are optimized together to minimize the Kullback-Leibler divergence between the true posterior distribution and the estimated one for every parameters of a given forward model. The ‘Neural Posterior Estimator’ (NPE) module [Papamakarios et al., 2017] uses normalizing flows [Papamakarios et al., 2021] to approximate the posterior distribution, while the ‘Multi-Layer Perceptron’ (MLP) module is used to reduce the data dimensionality and ensure fast and robust convergence of the NPE module.

μGUIDE framework.

μGUIDE takes as input an observed data vector and relies on the definition of a biophysical or computational model [Ascoli et al., 2007, Callaghan et al., 2020, Jelescu et al., 2020]. It outputs a posterior distribution of the model parameters. Based on a SBI framework, it combines a Multi-Layer Perceptron (MLP) with 3 layers and a Neural Posterior Estimator (NPE). The MLP learns a low-dimensional representation of x, based on a small number of features (Nf), that can be either defined a priori or determined empirically during training. The MLP is trained simultaneously with the NPE, leading to the extraction of the optimal features that minimize the bias and uncertainty of p(θ|x).

The full posterior distribution contains a lot of useful information. To summarize and easily visualize this information, we propose three measures that quantify the best estimates and the associated confidence levels, and a way to highlight degeneracy. The three measures are the Maximum A Posteriori (MAP), which corresponds to the most likely parameter estimate; an uncertainty measure, which quantifies the dispersion of the 50% most probable samples using the interquartile range, relative to the prior range; and an ambiguity measure, which measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range. Fig. 2 presents those measures on exemplar posterior distributions.

We show exemplar applications of μGUIDE to three biophysical models of increasing complexity and degeneracy from the dMRI literature: Ball&Stick [Behrens et al., 2003] (Model 1); Standard Model (SM) [Novikov et al., 2018a] (Model 2); and extended-SANDI [Palombo et al., 2020] (Model 3).

μGUIDE summarizes information contained in the estimated posterior distributions.

A) Examples of degenerate and non-degenerate posterior distributions. Two Gaussian distributions are fitted to the obtained posterior distribution, where the means and standard deviations are represented by the vertical lines and shaded areas. A voxel is considered as degenerate if the derivative of the fitted Gaussian distributions changes signs more than once (i.e. multiple local maxima), and if the two Gaussian distributions are not overlapping (the distance between the two Gaussian means is inferior to the sum of their standard deviations). B) Presentation of the measures introduced to quantify a posterior distribution on exemplar non-degenerate posterior distributions. Maximum A Posteriori (MAP): is the most likely parameter estimate (dashed vertical lines). Uncertainty: measures the dispersion of the 50% most probable samples using the interquartile range, with respect to the prior range. Ambiguity: measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range.

2.2 Evaluation of μGUIDE on simulations

2.2.1 Comparison with MCMC

We performed a comparison between the posterior distributions obtained using μGUIDE and MCMC, a classical Bayesian method. Fig. 3A shows posterior distributions on three exemplar simulations with SNR = 50 using the model 2 (SM), obtained with 15000 samples. Sharper and less biased posterior estimations are obtained using μGUIDE. Fig. 3B presents histograms for each model parameter of the bias between the ground truth value used to simulate a signal, and the MAP of the posterior distributions obtained with either μGUIDE or MCMC, on 200 simulations. Results indicate that the bias is similar or smaller using μGUIDE. Overall, μGUIDE posterior distributions are more accurate than the ones obtained with MCMC.

Moreover, it took on average 29.3 s to obtain the posterior distribution using MCMC on a GPU (NVIDIA GeForce GT 710) for one voxel, while it only took 0.02 s for μGUIDE. μGUIDE is about 1500 times faster than MCMC, which makes it more suitable for applying it on large datasets.

Comparison between μGUIDE and MCMC.

A) Posterior distributions obtained using either μGUIDE or MCMC on three exemplar simulations with model 2 (SM - SNR = 50). Names of the model parameters are indicated in the titles of the panels. B) Bias between the ground truth values used for simulating the diffusion signals, and the maximum-a-posteriori extracted from the posterior distributions using either μGUIDE or MCMC. Sharper and less biased posterior distributions are obtained using μGUIDE.

2.2.2 The importance of feature selection

Fig. 4 shows the MAP extracted from the posterior distributions versus the ground truth parameters used to generate the diffusion signal with μGUIDE and manually-defined summary statistics for the three models. Less biased MAPs with lower ambiguities and uncertainties are obtained with μGUIDE, indicating that the MLP allows for the extraction of additional information not contained in the summary statistics, helping to solve the inverse problem with higher accuracy and precision. μGUIDE generalizes the method developed in [Jallais et al., 2022] to make it applicable to any forward model and any acquisition protocol, while making the estimates more precise and accurate thanks to the automatic feature extraction.

2.2.3 μGUIDE highlights degeneracies

Fig. 5 presents the posterior distributions of microstructure parameters for the three models obtained with μGUIDE on exemplar noise-free simulations. Blue curves correspond to non-degenerate posterior distributions, while the red ones present at least one degeneracy for one of the parameters. As the complexity of the model increases, degeneracy in the model definitions appear. This figure showcases μGUIDE ability to highlight degeneracy in the model parameter estimation.

Tables 1 and 2 present the number of degenerate cases for each parameter in the three models, on 10000 simulations. Table 1 considers noise-free simulations and the training and estimations were performed on CPU. Table 2 reports results on noisy simulations (Rician noise with SNR = 50), with training and testing performed on a GPU (NVIDIA GeForce RTX 4090). The time needed for the inference and to estimate the posterior distributions on 10000 simulations, define if they are degenerate or not, and extract the MAP, uncertainty, ambiguity, are also reported. The more complex the model, the more degeneracies.

Fitting accuracy comparison between μGUIDE’s MLP-extracted features and manually-defined summary statistics.

Maximum-A-Posterioris extracted from the posterior distributions vs ground truth parameters used for generating the signal for the three models. Orange points correspond to the MAPs obtained using MLP-extracted features (μGUIDE) and the blue ones to the MAPs with the manually-defined summary statistics. Only the non-degenerate posterior distributions were kept. The summary statistics used in those three models are the direction-averaged signal for the Ball&Stick model, the LEMONADE system of equations [Novikov et al., 2018b] for the SM, and the summary statistics defined in [Jallais et al., 2022] for the extended-SANDI model. Results are shown on 100 exemplar noise-free simulations with random parameter combinations. The optimal features extracted by the MLP allow to reduce the bias and variance of the obtained microstructure posterior distributions.

2.3 Application of μGUIDE to real data

After demonstrating that the proposed framework provides good estimates in the controlled case of simulations, we applied μGUIDE to both a healthy volunteer and a participant with epilepsy. The estimation of the posterior distributions is done independently for each voxel. To easily assess the values and the quality of the fitting, we are plotting the maximum-a-posterior, ambiguity and uncertainty maps, but the full posterior distributions are stored and available for all the voxels. Voxels presenting a degeneracy are highlighted with a red dot.

2.3.1 Healthy volunteer

We applied μGUIDE to a healthy volunteer, using the Ball&Stick, SM and extended-SANDI models. Fig. 6 presents the parametric maps of an exemplar set of model parameters for each model, alongside their degeneracy, uncertainty and ambiguity. The Ball&Stick model presents no degeneracy, the SM presents some degeneracy, mostly in voxels with high likelihood of partial voluming with cerebrospinal fluid (CSF) and at the white matter - gray matter boundaries. The extended-SANDI model is the model showing the highest number of degenerate cases, mostly localized within the white matter areas characterized by complex microstructure, e.g. crossing fibers. This result is expected, as the complexity of the models increases, leading to more combinations of tissue parameters that can explain an observed signal. Measures of ambiguity and uncertainty allow to quantify the confidence in the estimates and help interpreting the results.

2.3.2 Participant with epilepsy

Fig. 7 demonstrates μGUIDE application to a participant with epilepsy, using the SM. Noteworthy, the axonal signal fraction estimates within the epileptic lesion show low uncertainty and ambiguity measures hence high confidence, while orientation dispersion index estimates show high uncertainty and ambiguity suggesting low confidence, cautioning the interpretation.

Exemplar posterior distributions of the microstructure parameters for the Ball&Stick, SM and extended SANDI models, obtained using μGUIDE on exemplar noise-free simulations.

As the complexity of the model increases, degeneracies appear (red posterior distributions). μGUIDE allows to highlight those degeneracies present in the model definition.

3 Discussion

3.1 Applicability of μGUIDE to multiple models

The μGUIDE framework offers the advantage of being easily applicable to various biophysical models and representations, thanks to its data-driven approach for data reduction. The need to manually define specific summary statistics that capture the relevant information for microstructure estimation from the multi-shell diffusion signal is removed. This also eliminates the acquisition constraints that were previously imposed by the summary statistics definition [Jallais et al., 2022]. The extracted features contain additional information compared to the summary statistics (see Appendix A), resulting in a notable reduction in bias (on average 5.2 fold lower), uncertainty (on average 2.6 fold lower) and ambiguity (on average 2.7 fold lower) in the estimated posterior distributions. Consequently, μGUIDE improves parameters estimation over current state-of-the-art methods (e.g. Jallais et al. [2022]), showing for example reduced bias (on average 5.2 fold lower) and dispersion (on average 6.4 fold lower) on the MAP estimates for each of the three example models investigated (see Fig. 4).

In this study, we presented applications of μGUIDE to brain microstructure estimation using three well-established biophysical models, with increased complexity: the Ball&Stick model, the Standard Model, and an extended-SANDI model. However, our approach is not limited to brain tissue nor to diffusion-weighted MRI and can be extended to different organs by employing their respective acquisition encoding and forward models, such as NEXI for exchange estimates [Jallais et al., 2024]; mcDESPOT for myelin water fraction mapping using quantitative MRI relaxation [Deoni et al., 2008]; VERDICT in prostate imaging [Panagiotaki et al., 2014]; or even adapted to different imaging modalities (e.g., EEG and MEG), where there is a way to link (via modelling or simulation) the observed signal to a set of parameters of interest. This versatility underscores the broad applicability of our proposed approach across various biological systems and imaging techniques.

It is important to note that μGUIDE is still a model-dependent method, meaning that the training process is based on the specific model being used. Additionally, the number of features extracted by the MLP needs to be predetermined. One way to determine the number of features is by matching it with the number of parameters being estimated. Alternatively, a dimensionality-reduction study using techniques like t-distributed stochastic neighbour embedding (tSNE) [Van der Maaten and Hinton, 2008] can be conducted to determine the optimal number of features.

3.2 μGUIDE: an efficient framework for Bayesian inference

One notable advantage of μGUIDE is its amortized nature. With this approach, the training process is performed only once, and thereafter, the posterior estimations can be independently obtained for all voxels. This amortization enables efficient estimations of the posterior distributions. μGUIDE outperforms in terms of speed conventional Bayesian inference methods such as MCMC, showing a ~1500 fold acceleration. The time savings achieved with μGUIDE make it a highly efficient and practical tool for estimating posterior distributions in a timely manner.

This unlocks the possibility to process with Bayesian inference very large datasets in manageable time (e.g. approximately 6 months to process 10k dMRI datasets) and to include Bayesian inference in iterative processes that require the repeated computation of the posterior distributions (e.g., dMRI acquisition optimization Alexander [2008]).

In the dMRI community, the use of SBI methods to characterize full posterior distributions as well as quantify the uncertainty in parameter estimations was first introduced in Jallais et al. [2022] for a grey matter model. An application to crossing fibers has recently been proposed by Karimi et al. [2024]. Those approaches use different density estimators. This work and Jallais et al. [2022] rely on Masked Autoregressive FLows (MAF) Papamakarios et al. [2017], while the work by Karimi et al. [2024] is based on Mixture Density Networks (MDN) [Bishop, 1994]. MAFs have been found to show superior performance compared to MDNs [Goncalves et al., 2020, Patron et al., 2022].

3.3 μGUIDE quantifies confidence to guide interpretation

Quantifying confidence in an estimate is of crucial importance. As demonstrated by our pathological example, changes in the tissue microstructure parameters can help clinicians decide which parameters are the most reliable and better interpret microstructure changes within diseased tissue. On large population studies, the quantified uncertainty can be taken into account when performing group statistics and to detect outliers.

Multiple approaches have been used to try and quantify this uncertainty. Gradient descent often provides a measure of confidence for each parameter estimate. Alternative approaches use the shape of the fitted tensor itself as a measure of uncertainty for the fiber direction [Koch et al., 2002, Parker and Alexander, 2003]. Other methods also rely on bootstrapping techniques to estimate uncertainty. Repetition bootstrapping for example depends on repeated measurements of signal for each gradient direction, but imply a long acquisition time and cost, and are prone to motion artifacts [Lazar and Alexander, 2005, Jones, 2003]. In contrast, residual bootstrapping methods resample the residuals of a regression model. Yet, this approach is heavily dependent on the model and can lead to overfitting [Whitcher et al., 2008, Chung et al., 2006]. In general, resampling methods can be problematic for sparse samples, as the bootstrapped samples tend to underestimate the true randomness of the distribution [Kauermann et al., 2009]. We propose to quantify the confidence by estimating full posterior distributions, which also has the benefit of highlighting degeneracy. Model-fitting methods with different initializations, as done in e.g. Jelescu et al. [2016], also allow to highlight degeneracies. However, they only provide a partial description of the solution landscape, which can be interpreted as a partial posterior distribution. In contrast, Bayesian methods estimate the full posterior distributions, offering a more accurate and precise characterization of degeneracies and uncertainties. Hence, in this work we decided to use MCMC, a traditional Bayesian method, as benchmark.

Parametric maps of the Ball&Stick (top), SM (middle) and extended-SANDI model (bottom), obtained using μGUIDE.

Maximum-a-posteriori estimation, uncertainty and ambiguity measure maps are reported, overlayed with voxels considered degenerate (red dots).

Parametric maps of a participant with epilepsy obtained using μGUIDE with the SM, superimposed with the grey matter (black) and white matter (white) lesions segmentation.

Mean values of the maximum-a-posterior, uncertainty and ambiguity measures are reported in the two regions of interest. Lower MAP values are obtained in the lesions for the axonal signal fraction and the orientation dispersion index compared to healthy tissue. Higher uncertainty and ambiguity ODI values are reported, suggesting less stable estimations.

Variance observed in the posterior distributions can be attributed to several factors. The presence of noise in the signal contributes to irreducible variance, decreasing the confidence in the estimates as the noise level increases (see Appendix B). Another source of variance can arise from the choice of acquisition parameters. Different acquisitions may provide varying levels of confidence in the parameter estimates. Under-sampled acquisitions or inadequate b-shells may fail to capture essential information about a tissue microstructure, such as soma or neurite radii, resulting in inaccurate estimates.

μGUIDE can guide users in determining whether an acquisition is suitable for estimating parameters of a given model and viceversa, the variance and bias of the posterior distributions estimated with μGUIDE can be used to guide the optimization of the data acquisition to maximize accuracy and precision of the model parameters estimates.

The presence of degeneracy in the solution of the inverse problem is influenced by the complexity of the model being used and the lack of sufficient information in the data. In recent years, researchers have introduced increasingly sophisticated models to better represent the brain tissue, such as SANDI [Palombo et al., 2020], NEXI [Jelescu et al., 2022] and eSANDIX [Olesen et al., 2022], that take into account an increasing number of tissue features. By applying μGUIDE, it becomes possible to gain insights into the degree of degeneracy within a model and to assess the balance between model realism and the ability to accurately invert the problem. We have recently provided an example of such application for NEXI and SANDIX [Jallais et al., 2024].

3.4 Summary

We propose a general Bayesian framework, dubbed μGUIDE, to efficiently estimate posterior distributions of tissue microstructure parameters. For any given acquisition and signal model/representation, μGUIDE improves parameters estimation and computational time over existing state-of-the-art methods. It allows to highlight degeneracy, and quantify confidence in the estimates, guiding results interpretation towards more confident and explainable diagnosis using modern deep learning. μGUIDE is not inherently limited to dMRI and microstructure imaging. We envision its usage and utility to perform efficient Bayesian inference also using data from any modality where there is a way to link (via modelling or simulation) the observed measurements to a set of parameters of interest.

4 Methods

4.1 Solving the inverse problem using Bayesian inference

4.1.1 The inference problem

We make the hypothesis that an observed dMRI signal x0 can be explained (and generated) using a handful of relevant tissue microstructure parameters θ0, following the definition of a forward model: x0=Μ(θ0)

The objective is, given this observation x0, to estimate the parameters θ0 that generated it.

Forward models are designed to mimic at best a given biophysical phenomenon, for some given time and scale [Alexander, 2009, Yablonskiy and Sukstanskii, 2010, Jelescu and Budde, 2017, Novikov et al., 2018a, Alexander et al., 2019, Jelescu et al., 2020]. As a consequence, forward models are injection functions (every biologically plausible θi generates exactly one signal xi), but do not always happen to be bijections, meaning that multiple θi can generate the same signal xi. It can be impossible, based on biological considerations, to infer which solution θi best reflects the probed structure. We refer to these models as ‘degenerate models’.

Point estimates algorithms, such as minimum least square or maximum likelihood estimation algorithms, allow to estimate one set of microstructure parameters that could explain an observed signal. In the case of degenerate models, the solution space can be multi-modal and those algorithms will hide possible solutions. When considering real-life acquisitions, i.e. noisy and/or under-sampled acquisitions, one also needs to consider the bias introduced with respect to the forward model, and the resulting variance in the estimates [Jones, 2003, Behrens et al., 2003].

We propose a new framework that allows for the estimation of full posterior distributions p(θ|x0), that is all the probable parameters that could represent the underlying tissue, along with an uncertainty measure and the interdependency of parameters. These posteriors can help interpreting the obtained results and make more informed decisions.

4.1.2 The Bayesian formalism

The posterior distribution can be defined using Bayes’ theorem as follows: p(θ|x0)=p(x0|θ)p(θ)p(x0), (1)

where p(x0|θ) is the likelihood of the observed data point, p(θ) is the prior distribution defining our initial knowledge of the parameter values, and p(x0) is a normalizing constant, commonly referred to as the evidence of the data.

The evidence term is usually very hard to estimate, as it corresponds to all the possible realisations of x0, i.e. p(x0) = ∫all x0p(x0|θ) p(θ) dx0. For simplification, methods usually estimate an unnormalized probability density function, i.e.

To approximate these posterior distributions, traditional methods rely on the estimation of the likelihood p(x0|θ) of the observed data point x0 via an analytic expression. This likelihood function corresponds to an integral over all possible trajectories through the latent space, that is p(x0|θ) = ∫ p(x0, z|θ)dz, where p(x0, z|θ) is the joint probability density of observed data x0 and latent variables z. For forward models with large latent spaces, computing this integral explicitly becomes impractical. The likelihood function is then intractable, rendering these methods unusable [Cranmer et al., 2020]. Models that do not admit a tractable likelihood are called implicit models [Diggle and Gratton, 1984].

To circumvent this issue, some techniques have been proposed to sample numerically from the likelihood function, such as Markov-Chain-Monte-Carlo (MCMC) [Metropolis et al., 1953]. Another set of approaches proposes to train a conditional density estimator to learn a surrogate of the likelihood distribution [Papamakarios et al., 2019, Lueckmann et al., 2019], the likelihoodratio [Cranmer et al., 2016, Gutmann et al., 2018] or the posterior distribution [Papamakarios and Murray, 2016, Lueckmann et al., 2017, Papamakarios et al., 2019], allowing to greatly reduce computation times. These methods are dubbed Likelihood-Free Inference (LFI), or SimulationBased Inference (SBI) methods [Cranmer et al., 2020, Tejero-Cantero et al., 2020]. In particular, there has been a growing interest towards deep generative modeling approaches in the machine learning community [Lueckmann et al., 2021]. They rely on specially tailored neural network architectures to approximate probability density functions from a set of examples. Normalizing flows [Papamakarios et al., 2021] are a particular class of such neural networks that have demonstrated promising results for SBI in different research fields [Goncalves et al., 2020, Greenberg et al., 2019].

While this work focuses on the estimate of the posterior distribution using a conditional density estimator, we show a comparison with MCMC, which are commonly used methods in the community. We will therefore introduce this method in the following paragraph.

4.1.3 Estimating the likelihood function

Well-established approaches for estimating the likelihood function are MCMC methods. These methods rely on a noise model to define the likelihood distribution, such as the Rician [Panagiotaki et al., 2012] or Offset Gaussian models [Alexander, 2009]. In this work, we will be using the Microstructure Diffusion Toolbox to perform the MCMC computations [Harms and Roebroeck, 2018], which relies on the Offset Gaussian model. The log-likelihood function is then the following: log(p(x|θ))=∑i=1m(xi−Μ(θ)i2+σ2)22σ2−m⋅log(σ2π), (3)

where M(θ) is the signal obtained using the biophysical model, M(θ)i is the ith measurement of the signal, σ is the standard deviation of the Gaussian distributed noise, estimated from the reconstructed magnitude images [Dietrich et al., 2007], and m is the number of observations in the dataset.

MCMC methods allow to obtain posterior distributions using Bayes’ formula (Eq. (2)) with the previously defined likelihood function (Eq. (3)) and some prior distributions, which are usually uniform distributions defined on biologically plausible ranges. They generate a multi-dimensional chain of samples which is guaranteed to converge towards a stationary distribution, which approximates the posterior distribution [Metropolis et al., 1953].

The need to compute the signal following the forward model at each iteration makes these sampling methods computationally expensive and time consuming. In addition, they require some adjustments specific to each model, such as the choice of burn-in length, thinning and the number of samples to store. Harms and Roebroeck [2018] recommend to use the Adaptive Metropolis-Within-Gibbs (AMWG) algorithm for sampling dMRI models, initialized with a maximum likelihood estimator obtained from non-linear optimization, with 100 to 200 samples for burn-in and no thinning. Authors notably investigated the use of starting from the MLE and thinning. They concluded that starting from the MLE allows to start in the stationary distribution of the Markov chain, and has the advantage of removing salt- and pepper-like noise from the resulting mean and standard deviation maps. Their findings also indicate that thinning is unnecessary and inefficient, and they recommend using more samples instead. The recommended number of samples is modeldependent. Authors recommendations can be found in their paper.

4.1.4 Bypassing the likelihood function

An alternative method was proposed to overcome the challenges associated with approximating the likelihood function and the limitations of MCMC sampling algorithms. This approach involves directly approximating the posterior distribution by using a conditional density estimator, i.e. a family of conditional probability density function approximators denoted as qϕ(θ|x). These approximators are parameterized by ϕ and accept both the parameters θ and the observation x as input arguments. Our posterior approximation is then obtained by minimizing its average Kullback-Leibler divergence with respect to the conditional density estimator for different choices of x, as per [Papamakarios and Murray, 2016]: minϕ. L(ϕ) with L(ϕ)=Ex∼p(x)[ DKL(p(θ|x)‖qϕ(θ|x)) ], (4)

which can be rewritten as ℒ(ϕ)=∫DKL(p(θ|x)||qϕ(θ|x))p(x)dx,=−∫∫log(qϕ(θ|x))p(θ|x)p(x)dθdx+C,=−∫∫log(qϕ(θ|x))p(x,θ)dθdx+C,=𝔼(x,θ)~p(x,θ)[ log(qϕ(θ|x)) ]+C, (5)

where C is a constant that does not depend on ϕ. Note that in practice we consider a N-sample Monte-Carlo approximation of the loss function: ℒ(ϕ)≈ℒN(ϕ)=−1N∑i=1Nlog(qϕ(θi|xi)), (6)

where the N data points (θi, xi) are sampled from the joint distribution with θi ~ p(θ) and xi ~ p(x|θi). We can then use stochastic gradient descent to obtain a set of parameters ϕ which minimizes LN.

If the class of conditional density estimators is sufficiently expressive, it can be demonstrated that the minimizer of Eq. (6) converges to p(θ|y) when N → ∞ [Greenberg et al., 2019]. It is worth noting that the parametrization ϕ, obtained at the end of the optimization procedure, serves as an amortized posterior for various choices of x. Hence, for a particular observation x0, we can simply use qϕ(θ|x0) as an approximation of p(θ|x0).

4.2 μGUIDE framework

The full architecture of the proposed Bayesian framework, dubbed μGUIDE, is presented in Fig. 1. The analysis codes underpinning the results presented here can be found upon publication in the Cardiff University data catalogue and on Github: https://github.com/mjallais/uGUIDE (both CPU and GPU are supported).

μGUIDE is comprised of two modules that are optimized together to minimize the Kull-back-Leibler divergence between the true posterior distribution and the estimated one for every parameters of a given forward model. The ‘Neural Posterior Estimator’ (NPE) module uses normalizing flows to approximate the posterior distribution, while the ‘Multi-Layer Perceptron’ (MLP) module is used to reduce the data dimensionality and ensure fast and robust convergence of the NPE module. The following Sections provide more details about our implementation of each module.

4.2.1 Neural Posterior Estimator

In this study, the Sequential Neural Posterior Estimation (SNPE-C) algorithm [Papamakarios and Murray, 2016, Greenberg et al., 2019] with a single round is employed to train a neural network that directly approximates the posterior distribution. Thus, sampling from the posterior can be done by sampling from the trained neural network. Neural density estimators have the advantage of providing exact density evaluations, in contrast to Variational Autoencoders (VAE) Kingma and Welling [2019] or generative adversarial networks (GAN) Goodfellow et al. [2014], which are better suited for generating synthetic data.

The conditional probability density function approximators used in this pro ject belong to a class of neural networks called normalizing flows [Papamakarios et al., 2021]. These flows are invertible functions capable of transforming vectors generated from a simple base distribution (e.g. the standard multivariate Gaussian distribution) into an approximation of the true posterior distribution. An autoregressive architecture for normalizing flows is employed, implemented via the Masked Autoregressive Flow (MAF) Papamakarios et al. [2017], which is constructed by stacking five Masked Autoencoder for Distribution Estimation models (MADE) Germain et al. [2015]. An explanation of how MAF and MADE work is provided in Appendix C.

To test that the predicted posteriors for a given model are not incorrect we use posterior predictive checks, which is described in more details in Appendix D.

4.2.2 Handling the large dimensionality of the data with Multi-Layer Perceptron

As the dimensionality of the input data x grows, the complexity of the corresponding inverse problem also increases. Accurately characterizing the posterior distributions or estimating the tissue microstructure parameters becomes more challenging. As a consequence, it is often necessary to rely on a set of low-dimensional features (or summary statistics) instead of the raw data for the inference task process [Blum et al., 2013, Fearnhead and Prangle, 2012, Papamakarios et al., 2019]. These summary statistics are features that capture the essential information within the raw data, allowing to reduce the size of the input vector. Learning a set of sufficient statistics before estimating the posterior distribution makes the inference easier and offers many benefits (see e.g. the Rao-Blackwell theorem).

A follow-up challenge lies in the choice of suitable summary statistics. For well-understood problems and data, it is possible to manually design these features using deterministic functions that condense the information contained in the raw signal into a set of handful summary statistics. Previous works, such as Novikov et al. [2018b] and Jallais et al. [2022], have proposed specific summary statistics for two different biophysical models. However, defining these summary statistics is difficult and often requires prior knowledge of the problem at hand. In the context of dMRI, they also rely on acquisition constraints and are model-specific.

In this work, the proposed framework aims to be applicable to any forward model and be as general as possible. We therefore propose to learn the summary statistics from the high-dimensional input signals x using a neural network. This neural network is referred to as an embedding neural network. The observed signals are fed into the embedding neural network, whose outputs are then passed to the neural density estimator. The parameters of the embedding network are learned together with the parameters of the neural density estimator, leading to the extraction of optimal features that minimize the uncertainty of p(θ|x). Here, we propose to use a MLP with three layers as a summary statistics extractor. The number of features Nf extracted by the MLP can be either defined a priori or determined empirically during training.

4.2.3 Training μGUIDE

To train μGUIDE we need couples of input vectors x and corresponding ground-truth values for the model parameters that we want to estimate, θ. The input x can be real or simulated data (e.g. dMRI signals); or a mixture of these two. We train μGUIDE by stochastically minimizing the loss function defined in Eq. (6) using the Adam optimizer [Kingma and Ba, 2015] with a learning rate of 10−3 and a minibatch size of 128. We use 1 million simulations for each model, 5% of which are randomly selected to be used as a validation set. Training is stopped when the validation loss does not decrease for 30 consecutive epochs.

4.2.4 Quantifying the confidence in the estimates

The full posterior distribution contains a lot of useful information about a given model parameter best estimates; uncertainty; ambiguity and degeneracy. To summarize and easily visualize this information, we propose three measures that quantify the best estimates and the associated confidence levels, and a way to highlight degeneracy.

We start by checking whether a posterior distribution is degenerate, that is if the distribution presents multiple distinct parameter solutions, appearing as multiple local maxima (Fig. 2). To that aim, we fit two Gaussian distributions to the obtained posterior distributions. A voxel is considered as degenerate if the derivative of the fitted Gaussian distributions changes signs more than once (i.e. multiple local maxima), and if the two Gaussian distributions are not overlapping (the distance between the two Gaussian means is inferior to the sum of their standard deviations).

For non-degenerate posterior distributions, we extract three quantities:

The Maximum A Posteriori (MAP), which corresponds to the most likely parameter estimate.

An uncertainty measure, which quantifies the dispersion of the 50% most probable samples using the interquartile range, relative to the prior range.

An ambiguity measure, which measures the Full Width at Half Maximum (FWHM), in percentage with respect to the prior range.

Fig. 2 presents those measures on exemplar posterior distributions.

4.3 Application of μGUIDE to biophysical modelling of dMRI data

We show exemplar applications of μGUIDE to three biophysical models of increasing complexity and degeneracy from the dMRI literature. For each model, we compare the fitting quality of the posterior distributions obtained using the MLP and manually defined summary statistics.

4.3.1 Biophysical models of dMRI signal.

Model 1: Ball&Stick [Behrens et al., 2003]. This is a two-compartment model (intra-neurite and extra-neurite space) where the dMRI signal from the brain tissue is modeled as a weighted sum, with weight fin, of signals from water diffusing inside the neurites, approximated as sticks (i.e. cylinders of zero radius) with diffusivity Din, and water diffusing within the extra-neurite space, approximated as Gaussian diffusion in an isotropic medium with diffusivity De. The direction of the stick is randomly sampled on a sphere. This model has the main advantage of being nondegenerate. We define the summary statistics as the direction-averaged signal (six b-shells, see Section 4.3.3).

Model 2: Standard Model (SM) [Novikov et al., 2018a]. Expanding on model 1, this model represents the dMRI signal from the brain tissue as a weighted sum of the signal from water diffusing within the neurite space, approximated as sticks with symmetric orientation dispersion following a Watson distribution and water diffusing within the extra-neurite space, modelled as anisotropic Gaussian diffusion. The microstructure parameters of this two-compartment model are the neurite signal fraction f , the intra-neurite diffusivity Da, the orientation dispersion index ODI, and the parallel and perpendicular diffusivities within the extra-neurite space De|| and De⊥ . We use the LEMONADE [Novikov et al., 2018b] system of equations, which is based on a cumulant decomposition of the signal, to define six summary statistics.

Model 3: extended-SANDI [Palombo et al., 2020]. This is a three-compartment model (intra-neurite, intra-soma and extra-cellular space) where the dMRI signal from the brain tissue is modelled as a weighted sum of the signal from water diffusing within the neurite space, approximated as sticks with symmetric orientation dispersion following a Watson distribution; water diffusing within cell bodies (namely soma), modelled as restricted diffusion in spheres; and water diffusing within the extra-cellular space, modelled as isotropic Gaussian diffusion. The parameters of interest are the neurite signal fraction fn, the intra-neurite diffusivity Dn, the orientation dispersion index ODI, the extra-cellular signal fraction fe and isotropic diffusivity De, the soma signal fraction fs, and a proxy of soma radius and diffusivity Cs, defined as [Jallais et al., 2022]: Cs=2Dsδ2∑m=1∞αm−4αm2rs2−2⋅(2δ−2+e−αm2Ds(Δ−δ)−e−αm2Dsδ−e−αm2DsΔ+e−αm2Ds(Δ+δ)αm2Ds),

with rs and Ds the soma radius and diffusivity respectively, and αm the mth root of (ars)−1 J32 (ars) = J52 (ars), with Jn(x) the Bessel functions of the first kind. We use the six summary statistics defined in [Jallais et al., 2022], which are based on a high and low b-value signal expansion. Signal fractions follow the rule fn + fs + fe = 1, leading to six parameters to estimate for this model.

Prior distributions p(θ) are defined as uniform distributions over biophysically plausible ranges. Signal fractions are defined within the interval [0, 1], diffusivities between 0.1 and 3 μm2 ms−1, ODI between 0.03 and 0.95, and Cs between 0.15 and 1105 μm2 (which correspond to rs ∈ [1; 15] pm and fixed Ds = 3 μm2 ms−1).

The SM imposes the constraint De⊥ < De|| . To generate samples uniformly distributed on the space defined by this condition, we are using two random variables u0 and u1, both sampled uniformly between 0 and 1, and then relate them to De|| and De⊥ using the following equations: { De||=(3.0−0.1)2⋅u0+0.1De⊥=(De||−0.1)⋅u1+0.1 (7)

The extended-SANDI model requires for the signal fractions to sum to 1, that is fn + fs + fe = 1. To uniformly cover the simplex fn + fs + fe = 1, we define two new parameters k1 and k2, uniformly sampled between 0 and 1, and use the following equations to get the corresponding signal fractions: { fn=k2k1fs=(1−k2)k1fe=1−k1 (8)

To ensure comparability of results, we extract the same number of features Nf using the MLP as the number of summary statistics for each model. We therefore use Nf = 6 for the Ball&Stick, the SM and the extended-SANDI models. Although the number of features predicted by the MLP is fixed to Nf = 6 for the three models, the characteristics of these six features can be very different, depending on the chosen forward model and the available data (see Appendix A). Training the MLP together with the NPE module allows to maximise inference performance in terms of accuracy and precision.

4.3.2 Validation in Numerical Simulations

We start by validating the proposed method using posterior predictive checks (PPC) and simulated signals from model 2 (see more details in Appendix D). Since PPC alone does not guarantee the correctness of the estimated posteriors, we further validated the obtained posterior distributions comparing them with the Adaptive Metropolis-Within-Gibbs MCMC [Roberts and Rosenthal, 2009]. We generated simulations following the same acquisition protocol as the real data (see Section 4.3.3), and added Gaussian noise to the real and imaginary parts of the simulated signal with a Signal-to-Noise-Ratio (SNR) of 50, and then used the magnitude of this noisy complex signal for our experiments. We then estimated the posterior distributions using both μGUIDE and the MCMC method implemented in the MDT toolbox [Harms and Roebroeck, 2018]. We initialized the sampling using a maximum likelihood estimator. We sampled 15200 samples from the distribution, the first 200 ones being used as burn-in, and no thinning. Similarly, we sampled 15000 samples from the estimated posterior distributions using μGUIDE.

Then, we show that μGUIDE can be applicable to any model. We use model 1 and 3 as examples of simpler (and non-degenerate) and more complex (and degenerate) models than model 2, respectively.

We compared the proposed framework to a state-of-the-art method for posterior estimation [Jallais et al., 2022]. This method relies on manually-defined summary statistics, while μGUIDE automatically extracts some features using an embedded neural network. μGUIDE was trained directly on noisy simulations. The manually-defined summary statistics were extracted from these simulated noisy signals and then used as training dataset for a MAF, similarly to [Jallais et al., 2022].

Finally, we used μGUIDE to highlight degeneracy in all the models. While the complexity of the models increases, more degeneracy can be found. The degeneracy is inherent to the model definition, and is not induced by the noise. μGUIDE allows to highlight those degeneracies and quantify the confidence in the obtained estimates.

The training was performed on N = 106 numerical simulations for each model, computed using the MISST package [lanus et al., 2017] and random combinations of the model parameters, each uniformly sampled from the previously defined ranges, with the addition of Rician distributed noise with SNR equivalent to the experimental data, i.e. 50.

4.3.3 dMRI Data Acquisition and Processing

We applied μGUIDE to dMRI data collected from two participants: a healthy volunteer from the WAND dataset [McNabb et al., 2024] and an age-matched participant with epilepsy, acquired with the same protocol used for the MICRA dataset [Koller et al., 2021]. Data were acquired on a Connectome 3T scanner using a single-shot spin-echo, echo-planar imaging sequence with b-values = [200, 500, 1200, 2400, 4000, 6000] s mm−2, [20, 20, 30, 61, 61, 61] uniformly distributed directions respectively and 13 non-diffusion-weighted images at 2 mm isotropic resolution. TR was set to 3000 ms, TE to 59 ms, and the diffusion gradient duration and separation to 7 ms and 24 ms respectively. Short diffusion times and TE were achieved thanks to the Connectom gradients, allowing to enhance the signal-to-noise ratio and sensitivity to small water displacements [Jones et al., 2018, Setsompop et al., 2013]. We considered the noise as Rician with an SNR of 50 for both subjects.

Data were preprocessed using a combination of in-house pipelines and tools from the FSL [Andersson et al., 2003, Andersson and Sotiropoulos, 2016, Smith, 2002, Smith et al., 2004] and MRTrix3 [Tournier et al., 2019] software packages. The preprocessing steps included brain extraction [Smith, 2002], denoising [Cordero-Grande et al., 2019, Veraart et al., 2016], drift correction [Vos et al., 2017, Sairanen et al., 2018], susceptibility-induced distortions [Andersson et al., 2003, Smith et al., 2004], motion and eddy current correction [Andersson and Sotiropoulos, 2016], correction for gradient non-linearity distortions [Glasser et al., 2013], and Gibbs ringing artefacts correction [Kellner et al., 2016].

4.3.4 dMRI Data Analysis

Diffusion signals were first normalized by the mean non-diffusion-weighted signals acquired for each voxel. Each voxel was then estimated in parallel using the μGUIDE framework. For each observed signal xv (i.e. for each voxel), we drew 50000 samples via rejection-sampling from qϕ(θi|xv) for each model parameter θi, allowing to retrieve the full posterior distributions. If a posterior distribution was not deemed degenerate, the maximum-a-posteriori, uncertainty and ambiguity measures were extracted from the posterior distributions.

The manually-defined summary statistics of the SM are defined using a cumulant expansion, which is only valid for small b-values. We therefore only used the b ≤ 2500 smm−2 data for this model. In order to obtain comparable results, we restricted the application of μGUIDE to this range of b-values as well. An extra b-shell (b-value = 5000 smm−2; 61 directions) was interpolated using mapl [Fick et al., 2016] for the extended-SANDI model when using the method developed by Jallais et al. [2022] based on summary statistics.

The training of μGUIDE was performed as described in 4.3.2 and an example of training dataset and input signal vector is provided in Fig. 8.

Example training set and input signals for μGUIDE.

A) Examples of input synthetic data vectors and corresponding ground truth model parameters used in the training set of Model 1 (Ball&Stick). B) Example of input measured signals from a voxel in a healthy participant, used for inference.

All the computations were performed both on CPU and GPU (NVIDIA GeForce RTX 4090).